---
title: "Statistical Analysis of Digital Access Patterns in Ethiopian Grade 12 Students"
author: "Robsen Shemsedin Yusuf"
date: "2026-02-12"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)
```

# Introduction

In this project, I analyze digital access patterns among Ethiopian Grade 12 students using anonymized data collected from my Android application, **Star Exams**, which I developed and published on the Google Play Store.

The dataset contains 5,203 anonymized user records. Each record includes non-personal technical and geographic information such as region, cleaned city name, smartphone brand, and device tier classification (**budget**, **midrange**, **upper_midrange**). No personal identifiers such as names, phone numbers, or passwords are included.

The goal of this analysis is to investigate regional disparities in smartphone access and evaluate whether device characteristics can predict a student's geographic region. Specifically, I aim to:

- Estimate and compare device tier proportions across Ethiopian regions.
- Test the association between region and device tier using chi-square inference.
- Fit logistic and multinomial regression models.
- Apply classification techniques such as LDA and QDA.
- Compare models using AIC, BIC, and cross-validation.
- Visualize patterns using barplots, confidence intervals, and confusion matrices.

# Data source and preparation

## Load required packages

```{r packages}
library(tidyverse)
library(forcats)
library(janitor)
library(nnet)
library(MASS)
library(caret)
```

## Load dataset (local first, otherwise GitHub)

```{r load-data}
data_url  <- "https://raw.githubusercontent.com/robsenshemsedin/star-exams-clean-dataset/main/ethiopia_student_device_dataset.csv"
local_csv <- "ethiopia_student_device_dataset.csv"

if (file.exists(local_csv)) {
  df_raw <- read.csv(local_csv, stringsAsFactors = FALSE)
} else {
  df_raw <- read.csv(data_url, stringsAsFactors = FALSE)
}

df <- janitor::clean_names(df_raw)
glimpse(df)
```

## Data cleaning and formatting

```{r cleaning}
df <- df %>%
  mutate(across(everything(), ~na_if(.x, ""))) %>%
  mutate(
    region = as.factor(region),
    device_brand = as.factor(device_brand),
    device_range = as.factor(device_range)
  ) %>%
  mutate(device_range = forcats::fct_relevel(device_range,
                                             "budget",
                                             "midrange",
                                             "upper_midrange"))

summary(df)
```

# Exploratory data analysis

## Overall distribution of device tiers

```{r tier-distribution}
tier_overall <- df %>%
  count(device_range) %>%
  mutate(proportion = n / sum(n))

tier_overall
```

**Interpretation:** This table shows the overall distribution of smartphone tiers used by students. A high proportion of budget devices suggests that many users rely on lower-end smartphones, which may affect the quality and reliability of digital learning.

```{r plot-tier}
ggplot(df, aes(x = device_range)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Overall Distribution of Smartphone Device Tiers",
    x = "Device Tier",
    y = "Number of Students"
  )
```

## Device tier distribution by region

```{r tier-region-table}
tier_region <- df %>%
  count(region, device_range) %>%
  group_by(region) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

tier_region
```

**Interpretation:** This table compares regions by the proportion of budget, midrange, and upper-midrange devices. Differences across regions can reflect unequal access to higher-quality devices.

```{r plot-tier-region}
ggplot(tier_region, aes(x = region, y = prop, fill = device_range)) +
  geom_col() +
  theme_minimal() +
  labs(
    title = "Device Tier Proportion by Region",
    x = "Region",
    y = "Proportion",
    fill = "Device Tier"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Confidence intervals for budget device share

I estimate 95% confidence intervals for the proportion of budget devices within each region.

```{r ci-budget}
budget_ci <- df %>%
  filter(!is.na(region), !is.na(device_range)) %>%
  group_by(region) %>%
  summarise(
    n_region = n(),
    x_budget = sum(device_range == "budget"),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    test = list(prop.test(x_budget, n_region)),
    p_hat = x_budget / n_region,
    conf_low = test$conf.int[1],
    conf_high = test$conf.int[2]
  ) %>%
  ungroup()

budget_ci
```

**Interpretation:** The confidence intervals show uncertainty around the estimated budget-device proportion in each region. Wider intervals usually occur in regions with smaller sample size.

```{r plot-ci}
ggplot(budget_ci, aes(x = reorder(region, p_hat), y = p_hat)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf_low, ymax = conf_high), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Budget Device Share by Region (95% CI)",
    x = "Region",
    y = "Estimated Proportion"
  )
```

# Chi-square test of association

I test whether device tier and region are statistically independent.

- **H0:** Device tier is independent of region.
- **H1:** Device tier distribution differs across regions.

```{r chi-square}
tab <- table(df$region, df$device_range)
tab

chi_test <- chisq.test(tab)
chi_test

min_expected <- min(chi_test$expected)
min_expected
```

**Interpretation:** The p-value is extremely small, so I reject H0 and conclude that device tier distribution varies across regions.  
Because the minimum expected count is small, I also compute a Monte Carlo p-value as a robustness check.

```{r chi-square-mc}
chi_mc <- chisq.test(tab, simulate.p.value = TRUE, B = 5000)
chi_mc
```

# Logistic regression (Addis Ababa vs other regions)

To simplify the classification problem, I group the outcome into **Addis** versus **Other**.

```{r region-group}
df <- df %>%
  mutate(
    region_group = ifelse(tolower(as.character(region)) == "addis ababa", "Addis", "Other"),
    region_group = as.factor(region_group)
  )

table(df$region_group)
```

```{r logistic}
d_log <- df %>%
  dplyr::select(region_group, device_range, device_brand) %>%
  tidyr::drop_na()

log_model <- glm(region_group ~ device_range + device_brand,
                 data = d_log,
                 family = binomial)

summary(log_model)

AIC(log_model)
BIC(log_model)
```

**Interpretation:** Logistic regression models the probability that a student is in **Addis** rather than **Other** given device tier and brand.  
AIC and BIC are used for comparing models (lower values indicate a better trade-off between fit and complexity).

# Multinomial logistic regression (predicting region)

I fit a multinomial regression model to predict **region** using device tier and brand.

```{r multinomial}
d_multi <- df %>%
  dplyr::select(region, device_range, device_brand) %>%
  tidyr::drop_na()

multi_model <- nnet::multinom(region ~ device_range + device_brand,
                              data = d_multi,
                              trace = FALSE)

summary(multi_model)

AIC(multi_model)
BIC(multi_model)
```

**Interpretation:** Multinomial regression extends logistic regression to multiple region categories. It helps evaluate whether device characteristics contain useful information for predicting geography.

# LDA and QDA classification (Addis vs other)

I apply LDA and QDA to classify **Addis vs Other** using a compact numeric representation of device tier.

## Train/test split

```{r split}
set.seed(123)

d_cls <- df %>%
  dplyr::select(region_group, device_range) %>%
  tidyr::drop_na() %>%
  mutate(
    region_group = droplevels(as.factor(region_group)),
    device_range = droplevels(as.factor(device_range))
  )

train_index <- caret::createDataPartition(d_cls$region_group, p = 0.8, list = FALSE)
train <- d_cls[train_index, ]
test  <- d_cls[-train_index, ]

train_num <- train %>%
  mutate(device_range_num = as.numeric(device_range)) %>%
  dplyr::select(region_group, device_range_num)

test_num <- test %>%
  mutate(device_range_num = as.numeric(device_range)) %>%
  dplyr::select(region_group, device_range_num)

table(train_num$region_group)
table(test_num$region_group)
```

## LDA

```{r lda}
lda_out <- tryCatch({
  lda_model <- MASS::lda(region_group ~ device_range_num, data = train_num)
  lda_pred <- predict(lda_model, newdata = test_num)$class
  caret::confusionMatrix(lda_pred, test_num$region_group)
}, error = function(e) {
  cat("LDA could not be fitted: ", e$message, "\n")
  NULL
})

lda_out
```

## QDA

```{r qda}
qda_out <- tryCatch({
  qda_model <- MASS::qda(region_group ~ device_range_num, data = train_num)
  qda_pred <- predict(qda_model, newdata = test_num)$class
  caret::confusionMatrix(qda_pred, test_num$region_group)
}, error = function(e) {
  cat("QDA could not be fitted: ", e$message, "\n")
  NULL
})

qda_out
```

**Interpretation:** The confusion matrices summarize classification performance on the test set. Accuracy measures overall correctness, while sensitivity/specificity show class-wise performance.

# Cross-validation (5-fold)

I estimate classification performance stability using 5-fold cross-validation.

```{r cv}
set.seed(123)

control <- trainControl(method = "cv", number = 5)

cv_lda <- caret::train(
  region_group ~ device_range,
  data = d_cls,
  method = "lda",
  trControl = control
)

cv_lda
```

**Interpretation:** Cross-validation evaluates the model on multiple splits of the data. This gives a more reliable estimate of performance than using a single train/test split.

# Conclusion

In this analysis, I studied digital access patterns among Ethiopian Grade 12 students using smartphone and regional data collected from my Star Exams application.

I found clear regional differences in device tier distribution, supported by confidence intervals and a chi-square test of association. I then evaluated how well device characteristics predict region using logistic regression, multinomial regression, and classification methods (LDA/QDA). Finally, I used cross-validation to assess classification stability.

Overall, the results provide evidence that smartphone access differs by region, which is consistent with regional disparities in digital learning readiness.